GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-04-21 manuscript.tex

\begin{itemize}
    \item new geometry serializations
    \begin{itemize}
        \item[$-$] GeoJSON, KML etc. Newer, popular formats
    \end{itemize} 
    \item new ontology classes to cater for more nuanced spatial information
    \item more spatial functions
    \begin{itemize}
        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item scalar spatial properties 
    \begin{itemize}
        \item[$-$] area, volume etc. alongside geometries
    \end{itemize} 
    \item better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item[$-$] potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item Internet protocol-based selection of different geometries for features
\end{itemize}

@@

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
many 1.1 requests which are the only ones we report here. The next sections foreshadow likely 1.2 and 2.0 updates.

@@

In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0 
for use within SPARQL \texttt{FILTER} operations, 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.
% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

@@

>>>>

\begin{itemize}
    \item new geometry serializations
    \begin{itemize}
        \item[$-$] GeoJSON, KML and other popular formats
    \end{itemize} 
    \item new ontology classes to cater for more nuanced spatial information
    \item more spatial functions
    \begin{itemize}
        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item scalar spatial properties 
    \begin{itemize}
        \item[$-$] area, volume etc. alongside geometries
    \end{itemize} 
    \item better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item[$-$] potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item Internet protocol-based selection of different geometries for features
\end{itemize}

@@

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of April, 2021) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
many 1.1 requests which are the only ones we report here and which are accessible through the living document for the GeoSPARQL 1.1 standard \cite{ogcgeosparql11}. The next sections foreshadow likely 1.2 and 2.0 updates.

@@

In addition to \href{http://www.opengis.net/def/function/geosparql/union}{\emph{geof:union}}, \href{http://www.opengis.net/def/function/geosparql/envelope}{\emph{geof:envelope}} and \href{http://www.opengis.net/def/function/geosparql/convexHull}{\emph{geof:convexHull}} defined in GeoSPARQL 1.0 
for use within SPARQL \texttt{FILTER} operations, 1.1 defines \href{http://www.opengis.net/def/function/geosparql/union2}{\emph{geof:union2}} and 
as well as \href{http://www.opengis.net/def/function/geosparql/boundingCircle}{\emph{geof:boundingCircle}}, \href{http://www.opengis.net/def/function/geosparql/centroid}{\emph{geof:centroid}}, \href{http://www.opengis.net/def/function/geosparql/concatLines}{\emph{geof:ConcatLines}} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.

% \todo[inline]{Timo: I will create a pull request in the GeoSPARQL repository to clear up the issues concerning the spatial aggregate functions. Then I will update the namespaces here. I would find it nice to also include the functions we have discussed yesterday: transformation of SRS and in between literal formats. Should we mention it here as future work or should we make a draft pull request for that so that people have a reference? I think the latter}
\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geof:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

@@

In parallel, the OGC API Features standard \cite{ogcapifeaturespart3} is developed, which also proposes functions for filtering feature collections. While OGC API Features proposes the use of the Common Query Language (CQL) for filtering, it is also open to other query language implementations such as GeoSPARQL. When comparing the filter capabilities of CQL to GeoSPARQL, one can observe that concerning spatial filter capabilities, the two query languages provide comparable functionality. However, the version of CQL proposed by OGC API Features also provides support of spatiotemporal operators, which may be an addition to be further explored in the contiuous GeoSPARQL development process.
